---
title: "prosjekt3"
author: "Silje Anfindsen"
date: "11/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages(glmmTMB)
```

##  Problem 2

```{r load data }
long <- read.csv("https://www.math.ntnu.no/emner/TMA4315/2020h/eliteserie.csv", colClasses = c("factor","factor","factor","numeric"))
```

### 2 a) 
```{r fit model}
library(glmmTMB)
mod <- glmmTMB(goals ~ home + (1|attack) + (1|defence), poisson, data=long, REML=TRUE)
```

#### Part 1
State the precise assumptions of this model in suitable mathematical notation.

#### Part 2
Explain why the Poisson assumption may be reasonable way to model the inherent randomness of a football game.

Poisson distribution can be used to measure the probability of independent events occurring a certain number of times within a set period - such as the number of goals scored in a football match.

It can be used to do this by converting averages into a probability for the changeable outcomes.

The number of events occuring within a time interval or a region, is independent of the number of events that occurs in any other disjoint (non-overlapping) time interval or region.
The probability that a single event occurs within a small time interval or region, is proportional to the length of the interval or the size of the region.
The probability that more than one event may occur within a small time interval or region is negligable

When all of these three properties are funfilled we have a Poisson process.
The number of events in a Poisson process follows a Poisson distribution.

https://help.smarkets.com/hc/en-gb/articles/115001457989-How-to-calculate-Poisson-distribution-for-football-betting

### 2b

```{r summary}
summary(mod)
ranef(mod)
```

#### Part 1
Briefly discuss if the various parameter estimates appear reasonable.